---
title: "COI genetic diversity"
output: html_notebook
---

```{r message=FALSE}
pacman:: p_load("ape","pegas","adegenet","haplotypes","mmod",
                "tidyverse","readxl","ggplot2","ggpubr","ggrepel","gridExtra",
                "FSA","outliers","car","robustbase","MuMIn","rstatix")
```

# Import fasta

```{r}
seq <- read.dna("../Data/COI.fasta", format = "fasta")
seq
```

## population names

-   extract population names from sequence name
-   how many samples of each population

```{r}
# population name of each sequence
pop <- strsplit(dimnames(seq)[[1]], split="_") %>% sapply(function(x) x[4])
```

## sampling sites

```{r}
site <- read.csv("../Data/sites.csv", header = TRUE, sep = ",")
site
```

# calculate genetic diversity

## nucleotide diversity

-   calculate nucleotide diversity, and export in data frame, sort by nucleotide diversity

```{r eval=FALSE}
df.div <- data.frame(pop = levels(as.factor(pop)), nu.div = c(
  nuc.div(seq[1:10,]),     # BK
  nuc.div(seq[11:19,]),    # BTP
  nuc.div(seq[20:40,]),    # CLS
  nuc.div(seq[41:53,]),    # DFP
  nuc.div(seq[54:72,]),    # DHP
  nuc.div(seq[73:92,]),    # DK1-4
  nuc.div(seq[93:110,]),   # DK7-8
  nuc.div(seq[111:137,]),  # DKP
  nuc.div(seq[138:147,]),  # DTZ
  nuc.div(seq[148:155,]),  # FLP
  nuc.div(seq[156:167,]),  # JJP
  nuc.div(seq[168:180,]),  # NMNS
  nuc.div(seq[181:190,]),  # PLS
  nuc.div(seq[191:218,]),  # THU
  nuc.div(seq[219:227,]),  # WFC
  nuc.div(seq[228:238,]),  # XLC
  nuc.div(seq[239:245,]))) # ZGZ

# sort by nucleotide diversity
df.div[order(df.div$nu.div),]
```

## halotype diversity

-   calculate haplotype diversity, and export in data frame, sort by haplotype diversity

```{r eval=FALSE}
df.hap <- data.frame(pop = levels(as.factor(pop)), hap.div = c(
  hap.div(seq[1:10,]),     # BK
  hap.div(seq[11:19,]),    # BTP
  hap.div(seq[20:40,]),    # CLS
  hap.div(seq[41:53,]),    # DFP
  hap.div(seq[54:72,]),    # DHP
  hap.div(seq[73:92,]),    # DK1-4
  hap.div(seq[93:110,]),   # DK7-8
  hap.div(seq[111:137,]),  # DKP
  hap.div(seq[138:147,]),  # DTZ
  hap.div(seq[148:155,]),  # FLP
  hap.div(seq[156:167,]),  # JJP
  hap.div(seq[168:180,]),  # NMNS
  hap.div(seq[181:190,]),  # PLS
  hap.div(seq[191:218,]),  # THU
  hap.div(seq[219:227,]),  # WFC
  hap.div(seq[228:238,]),  # XLC
  hap.div(seq[239:245,]))) # ZGZ

# sort by nucleotide diversity
df.hap[order(df.hap$hap.div),]
```

## combine genetic diversity & sampling sites

-   merge site and genetic diversity data

```{r}
coi <- merge(site, df.div, by = "pop") %>% merge(df.hap, by = "pop")
coi
```

# Nucleotide diversity \~ PC1 score

-   Normality & outliers

```{r}
shapiro.test(coi$nu.div)
outlier(coi$nu.div)
```

## linear model (landscape)

```{r}
nu.lands.fit <- lm(nu.div ~ pc1_lands, data = coi)
summary(nu.lands.fit)
```

-   Diagnostics

```{r}
plot(nu.lands.fit)
```

-   normality & homoscendasticity

```{r}
# shapiro-wilk
residuals(nu.lands.fit) %>% shapiro.test()

# Bartlett
ncvTest(nu.lands.fit)

# Cook's distance
cooks.distance(nu.lands.fit) %>% plot()
which(cooks.distance(nu.lands.fit) > 4/length(coi$pop))
```

### robust regression model (landscape)

```{r}
nu.lands.rob <- lmrob(nu.div ~ pc1_lands, data = coi)
summary(nu.lands.rob)
```

## EXPORT-- plot (landscape)

```{r}
p1 <- 
ggplot(data = coi, aes(x = pc1_lands, y = nu.div)) + geom_point(size = 5) +
  geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) +
  geom_smooth(method = "lmrob", formula = y ~ x, color = "lightcoral", se = TRUE, fill = "lightcoral") +
  # geom_abline(intercept = coef(nu.lands.rob)[1], slope = coef(nu.lands.rob)[2], color = "black") +
  # annotate("text", x = -2, y = 0.005, label = "p = 0.86", hjust = 0, size = 8) +
  # annotate("text", x = -2, y = 0.005, label = "italic(R[adj])^2 == -0.06",
  #          parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
  theme_classic() +
  xlab("PC1") + ylab(expression(nucleotide~diversity~(pi))) + ggtitle("(A)") +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
        plot.title = element_text(size = 25, face = "bold"))

print(p1)
```

## linear model (local)

```{r}
nu.local.fit <- lm(nu.div ~ pc1_local, data = coi)
summary(nu.local.fit)
```

-   Diagnostics

```{r}
plot(nu.local.fit)
```

-   normality & homoscendasticity

```{r}
# shapiro-wilk
residuals(nu.local.fit) %>% shapiro.test()

# Bartlett
ncvTest(nu.local.fit)

# Cook's distance
cooks.distance(nu.local.fit) %>% plot()
which(cooks.distance(nu.local.fit) > 4/length(coi$pop))
```

### robust regression model (local)

```{r}
nu.local.rob <- lmrob(nu.div ~ pc1_local, data = coi)
summary(nu.local.rob)
```

## EXPORT-- plot (local)

```{r}
p2 <- 
ggplot(data = coi, aes(x = pc1_local, y = nu.div)) + geom_point(size = 5) +
  geom_smooth(method = "lm", formula = y ~ x,color = "cornflowerblue", se = TRUE) + 
  geom_smooth(method = "lmrob", formula = y ~ x, color = "lightcoral", se = TRUE, fill = "lightcoral") +
  # annotate("text", x = -2, y = 0.005, label = "p = 0.70", hjust = 0, size = 8) +
  # annotate("text", x = -2, y = 0.005, label = "italic(R[adj])^2 == -0.06",
  #          parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
  theme_classic() +
  xlab("PC1") + ylab(expression(nucleotide~diversity~(pi))) + ggtitle("(B)") +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
        plot.title = element_text(size = 25, face = "bold"))

print(p2)
```

# Haplotype diversity \~ PC1

-   Normality & outliers

```{r}
shapiro.test(coi$hap.div)
outlier(coi$hap.div)
```

## linear model (landscape)

```{r}
hap.lands.fit <- lm(hap.div ~ pc1_lands, data = coi)
summary(hap.lands.fit)
```

-   Diagnostics

```{r}
plot(hap.lands.fit)
```

-   normality & homoscendasticity

```{r}
# shapiro-wilk
residuals(hap.lands.fit) %>% shapiro.test()

# Bartlett
ncvTest(hap.lands.fit)

# Cook's distance
cooks.distance(hap.lands.fit) %>% plot()
which(cooks.distance(hap.lands.fit) > 4/length(coi$pop))
```

## EXPOET -- plot (landscape)

```{r}
p3 <- 
ggplot(data = coi, aes(x = pc1_lands, y = hap.div)) + geom_point(size = 5) +
  geom_smooth(method = "lm", color = "cornflowerblue", se = TRUE) + 
  # annotate("text", x = 2, y = 1, label = "p = 0.18", hjust = 0, size = 8) +
  # annotate("text", x = 2, y = 1, label = "italic(R[adj])^2==0.06", 
  #          parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
  theme_classic() +
  xlab("PC1") + ylab("Haplotype diversity (h)") + ggtitle("(C)") +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
        plot.title = element_text(size = 25, face = "bold"))

print(p3)
```

## linear model (local)

```{r}
hap.local.fit <- lm(hap.div ~ pc1_local, data = coi)
summary(hap.local.fit)
```

-   Diagnostics

```{r}
plot(hap.local.fit)
```

-   normality & homoscendasticity

```{r}
# shapiro-wilk
residuals(hap.local.fit) %>% shapiro.test()

# Bartlett
ncvTest(hap.local.fit)

# Cook's distance
cooks.distance(hap.local.fit) %>% plot()
which(cooks.distance(hap.local.fit) > 4/length(coi$pop))
```

## EXPORT -- plot (local)

```{r}
p4 <- 
ggplot(data = coi, aes(x = pc1_local, y = hap.div)) + geom_point(size = 5) +
  geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) + 
  # annotate("text", x = 1.2, y = 1, label = "p = 0.13", hjust = 0, size = 8) +
  # annotate("text", x = 1.2, y = 1, label = "italic(R[adj])^2==0.08", 
  #          parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
  theme_classic() +
  xlab("PC1") + ylab("Haplotype diversity (h)") + ggtitle("(D)") +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
        plot.title = element_text(size = 25, face = "bold"))

print(p4)
```

## EXPORT -- nucleotide & haplotype diversity

```{r}
fig3 <- grid.arrange(p1, p2, p3, p4, ncol = 2)
```

-   tif

```{r eval=FALSE}
ggsave(filename = "Figure3.tif", plot = fig3, width = 18, height = 12, units = "in", dpi = 150)
```

------------------------------------------------------------------------

# Diversity & Land use

## import landuse data

```{r}
landuse.lands <- read.csv("../Data/Landuse1000.csv")
landuse.local <- read.csv("../Data/Landuse250.csv")
```

- convert m^2 to km^2
```{r}
landuse.lands <- landuse.lands %>% mutate(across(2:9, ~ .x / 1000000))
landuse.local <- landuse.local %>% mutate(across(2:9, ~ .x / 1000000))

# check area
landuse.lands[ ,2:9] %>% rowSums()
landuse.local[ ,2:9] %>% rowSums()
```


- combine land use & diversity data

```{r}
landuse.lands <- landuse.lands %>% merge(coi[ ,c(1,7:8)], by = "pop")
landuse.local <- landuse.local %>% merge(coi[ ,c(1,7:8)], by = "pop")
landuse.lands; landuse.local
```


## Nucleotide diversity (landscape)

### each land use

```{r eval=FALSE}
# nucleotide diversity
lm(nu.div ~ Agriculture, data = landuse.lands) %>% summary()
lm(nu.div ~ Forest, data = landuse.lands) %>% summary()
lm(nu.div ~ Travel, data = landuse.lands) %>% summary()
lm(nu.div ~ Water, data = landuse.lands) %>% summary()
lm(nu.div ~ Build, data = landuse.lands) %>% summary()
lm(nu.div ~ Park, data = landuse.lands) %>% summary()
lm(nu.div ~ Grass, data = landuse.lands) %>% summary()
lm(nu.div ~ Other, data = landuse.lands) %>% summary()
```

### model selection (MuMIn)
- full model
```{r}
nu.lands.full <- lm(nu.div ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.lands)
summary(nu.lands.full)
```

```{r}
options(na.action = "na.fail")
dredge(nu.lands.full) %>% subset(delta < 1)
```

-   Best model

```{r}
dredge(nu.lands.full) %>% get.models(1) %>% .[[1]] %>% summary()
```

- VIF
```{r}
vif(lm(formula = nu.div ~ Forest + Grass + Other + Travel, data = landuse.lands))
```

-   Diagnostics

```{r}
plot(lm(formula = nu.div ~ Forest + Grass + Other + Travel, data = landuse.lands))
```

-   normality & homoscendasticity

```{r}
# shapiro-wilk
residuals(lm(formula = nu.div ~ Forest + Grass + Other + Travel, data = landuse.lands)) %>% shapiro.test()

# Bartlett
ncvTest(lm(formula = nu.div ~ Forest + Grass + Other + Travel, data = landuse.lands))

# Cook's distance
cooks.distance(lm(formula = nu.div ~ Forest + Grass + Other + Travel, data = landuse.lands)) %>% plot()
which(cooks.distance(lm(formula = nu.div ~ Forest + Grass + Other + Travel, data = landuse.lands)) > 4/length(coi$pop))
```

## Nucleotide diversity (local)
 
### each land use

```{r eval=FALSE}
# nucleotide diversity
lm(nu.div ~ Agriculture, data = landuse.local) %>% summary()
lm(nu.div ~ Forest, data = landuse.local) %>% summary()
lm(nu.div ~ Travel, data = landuse.local) %>% summary()
lm(nu.div ~ Water, data = landuse.local) %>% summary()
lm(nu.div ~ Build, data = landuse.local) %>% summary()
lm(nu.div ~ Park, data = landuse.local) %>% summary()
lm(nu.div ~ Grass, data = landuse.local) %>% summary()
lm(nu.div ~ Other, data = landuse.local) %>% summary()
```

### model selection (MuMIn)
- full model
```{r}
nu.local.full <- lm(nu.div ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.local)
summary(nu.local.full)
```

```{r}
options(na.action = "na.fail")
dredge(nu.local.full) %>% subset(delta < 1)
```

-   Best model

```{r}
dredge(nu.local.full) %>% get.models(1) %>% .[[1]] %>% summary()
```



## Haplotype diversity (landscape)
### each land use

```{r eval=FALSE}
lm(hap.div ~ Agriculture, data = landuse.lands) %>% summary()
lm(hap.div ~ Forest, data = landuse.lands) %>% summary()
lm(hap.div ~ Travel, data = landuse.lands) %>% summary()
lm(hap.div ~ Water, data = landuse.lands) %>% summary()
lm(hap.div ~ Build, data = landuse.lands) %>% summary()
lm(hap.div ~ Park, data = landuse.lands) %>% summary()
lm(hap.div ~ Grass, data = landuse.lands) %>% summary()
lm(hap.div ~ Other, data = landuse.lands) %>% summary()
```

### model selection (MuMIn)
- full model
```{r}
hap.lands.full <- lm(hap.div ~ Agriculture + Forest + Travel + Water +
                     Build + Park + Grass + Other, data = landuse.lands)
summary(hap.lands.full)
```

```{r}
options(na.action = "na.fail")
dredge(hap.lands.full) %>% subset(delta < 1)
```

-   Best model

```{r}
dredge(hap.lands.full) %>% get.models(1) %>% .[[1]] %>% summary()
```

- VIF
```{r}
vif(lm(formula = hap.div ~ Agriculture + Forest + Grass, data = landuse.lands))
```

- Diagnostics

```{r}
plot(lm(formula = hap.div ~ Agriculture + Forest + Grass, data = landuse.lands))
```

-   normality & homoscendasticity

```{r}
# shapiro-wilk
residuals(lm(formula = hap.div ~ Agriculture + Forest + Grass, data = landuse.lands)) %>% shapiro.test()

# Bartlett
ncvTest(lm(formula = hap.div ~ Agriculture + Forest + Grass, data = landuse.lands))

# Cook's distance
cooks.distance(lm(formula = hap.div ~ Agriculture + Forest + Grass, data = landuse.lands)) %>% plot()
which(cooks.distance(lm(formula = hap.div ~ Agriculture + Forest + Grass, data = landuse.lands)) > 4/length(coi$pop))
```

## Haplotype diversity (local)
### each land use

```{r eval=FALSE}
lm(hap.div ~ Agriculture, data = landuse.local) %>% summary()
lm(hap.div ~ Forest, data = landuse.local) %>% summary()
lm(hap.div ~ Travel, data = landuse.local) %>% summary()
lm(hap.div ~ Water, data = landuse.local) %>% summary()
lm(hap.div ~ Build, data = landuse.local) %>% summary()
lm(hap.div ~ Park, data = landuse.local) %>% summary()
lm(hap.div ~ Grass, data = landuse.local) %>% summary()
lm(hap.div ~ Other, data = landuse.local) %>% summary()
```

### model selection (MuMIn)
- full model
```{r}
hap.local.full <- lm(hap.div ~ Agriculture + Forest + Travel + Water +
                     Build + Park + Grass + Other, data = landuse.local)
summary(hap.local.full)
```

```{r}
options(na.action = "na.fail")
dredge(hap.local.full) %>% subset(delta < 1)
```

-   Best model

```{r}
dredge(hap.local.full) %>% get.models(1) %>% .[[1]] %>% summary()
```



----------------------------------------------------------------------------------------------------
# Genetic differentiation (FST)
## convert DNAbin to genind
```{r}
gen <- DNAbin2genind(seq)
gen
```

-   population strata
```{r}
strata(gen) <- data.frame(population = pop)
setPop(gen) <- ~population
gen
```

## Pairwise population differentiation
```{r eval=FALSE}
pfst <- pairwise_Gst_Nei(gen)
```

```{r eval=FALSE}
bs <- chao_bootstrap(gen, nreps = 100)
summarise_bootstrap(bs, D_Jost)
```

- convert dist to matrix
```{r}
pfst.mat <- as.matrix(pfst)
pfst.mat[upper.tri(pfst.mat, diag = TRUE)] <- NA
```

-   convert matrix to two column dataframe
-   negative values as zero
```{r}
fst.df <- pfst.mat %>% as.table() %>% as.data.frame() %>% filter(Freq != "NA")
colnames(fst.df) <- c("pop1", "pop2", "Fst")
fst.df$Fst[fst.df$Fst < 0] <- 0
head(fst.df)
```

-   urbanization region

```{r}
fst.df <- fst.df %>% mutate(region1 = case_when(
                            pop1 %in% c('FLP','NMNS','THU','DFP','DHP','ZGZ') ~ "Urban",
                            pop1 %in% c('BTP','XLC','DKP','DTZ','PLS','DK7-8') ~ "Suburban",
                            pop1 %in% c('WFC','JJF','CLS','DK1-4','BK') ~ "Nonurban")) %>% 

                     mutate(region2 = case_when(
                            pop2 %in% c('FLP','NMNS','THU','DFP','DHP','ZGZ') ~ "Urban",
                            pop2 %in% c('BTP','XLC','DKP','DTZ','PLS','DK7-8') ~ "Suburban",
                            pop2 %in% c('WFC','JJF','CLS','DK1-4','BK') ~ "Nonurban"))

head(fst.df)
```

```{r}
fst.df$pair <- paste(fst.df$region1, fst.df$region2, sep = "–")
levels(as.factor(fst.df$pair))
```

-   some pairs are identical
```{r}
fst.df$pair <- ifelse(fst.df$pair == "Nonurban–Urban", "Urban–Nonurban",
                      ifelse(fst.df$pair == "Nonurban–Suburban", "Suburban–Nonurban",
                             ifelse(fst.df$pair == "Suburban–Urban", "Urban–Suburban", fst.df$pair)))
```

-   reorder pairs
```{r}
fst.df$pair <- factor(fst.df$pair, levels = c("Urban–Urban","Suburban–Suburban","Nonurban–Nonurban",
                                              "Urban–Suburban","Urban–Nonurban","Suburban–Nonurban"))
levels(as.factor(fst.df$pair))    # check levels
```

## statistics
```{r}
# kruskal test
fst.df[fst.df$pair %in% c("Urban–Urban","Suburban–Suburban","Nonurban–Nonurban"),] %>% 
   kruskal.test(Fst ~ pair, data = .)   # p-value = 0.03174
```

-   Dunn's test
```{r}
fst.df[fst.df$pair %in% c("Urban–Urban","Suburban–Suburban","Nonurban–Nonurban"),] %>% 
  dunn_test(Fst ~ pair, p.adjust.method = "bonferroni")
```

## summarize

```{r}
fst.sum <- Summarize(Fst ~ pair, data = fst.df)
fst.sum
```

## EXPORT (dot plot)
```{r}
fst.plot <- fst.sum[c(1:3),] %>% mutate(pair = c("Urban", "Suburban", "Non-urban"))
fst.plot$pair <- factor(fst.plot$pair, levels = c("Urban", "Suburban", "Non-urban"))
fst.plot
```

- dot plot
```{r}
p.fst <- 
ggplot(data = fst.plot) + 
       geom_pointrange(aes(x = pair, y = mean, ymin = mean-sd/sqrt(n), ymax = mean+sd/sqrt(n),
                           color = pair), size = 1.5) +
       scale_color_manual(values = c("hotpink2","cornflowerblue","mediumseagreen")) +
       labs(x = "", color = "Pair") +  ylab(expression(Pairwise~~F[ST])) +
       theme_classic() +
       theme(axis.title = element_text(size = 15), axis.text.x = element_text(size = 15),
             plot.title = element_text(size = 20, face = "bold")) + 
       theme(legend.position = "none") +
       scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
       geom_segment(data = fst.plot, aes(x = "Urban", xend = "Non-urban", y = 0.07, yend = 0.07)) +
       geom_text(data = fst.plot, aes(x = "Suburban", y = 0.065), label = "*", size = 10) +
       ggtitle("(A)")
p.fst
```


