---
title: "SNP genetic diversity"
output: html_notebook
---

```{r message=FALSE}
library(adegenet)
library(hierfstat)
library(genepop)

library(readxl)
library(tidyverse)
library(graph4lg)
library(ggplot2)
library(gridExtra)

library(outliers)
library(MuMIn)
library(car)
library(rstatix)
library(FSA)
```

# import snp data
```{r eval=FALSE}
snp <- read.genepop("../Data/snp.gen", ncode = 3)  # adegenet package
snp
```

-   pouplations
```{r}
pop <- strsplit(rownames(snp@tab), split = "_") %>% sapply(function(x) x[1])
snp@pop <- as.factor(pop)   # population names
snp
```

- sample size
```{r}
summary(snp$pop)
```

## create hierfstat object
```{r eval=FALSE} 
fstat <- genind2hierfstat(snp)
fstat[1:5,1:5]
```

# calculate genetic diversity
```{r eval=FALSE}
div <- basic.stats(fstat, diploid = TRUE, digits = 3)
summary(div)
```

## Expected Heterozygosity (Hs)
```{r}
hs <- apply(div$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE)
hs
```

## Observed Heterozygosity (Ho)
```{r}
ho <- apply(div$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE)
ho
```

## Allelic richness (Ar)
```{r eval=FALSE}
ArRic <- allelic.richness(fstat)
```

-   mean allelic richness per population

```{r}
Ar <- ArRic$Ar %>% apply(MARGIN = 2, FUN = mean, na.rm = TRUE)
Ar
```

- combined
```{r}
div.df <- data.frame(He = hs, Ho = ho, Ar = Ar) %>% rownames_to_column(var = "pop")
div.df
```

# sampling sites
```{r}
site <- read.csv("../Data/sites.csv", header = TRUE, sep = ",")
site
```

## combine genetic diversity & sampling sites
-   merge site and genetic diversity data

```{r}
div.df <- merge(site, div.df, by = "pop")
```


# Genetic diversity ~ PC1
- Normality
```{r}
shapiro.test(div.df$He)
shapiro.test(div.df$Ho)
shapiro.test(div.df$Ar)
```

- outliers
```{r}
outlier(div.df$He)
outlier(div.df$Ho)
outlier(div.df$Ar)
```

## He ~ PC1
### linear model (landscape)
```{r}
he.lands.fit <- lm(He ~ pc1_lands, data = div.df)
summary(he.lands.fit)
```

-   Diagnostics
```{r}
plot(he.lands.fit)
```

- normality & homoscendasticity
```{r}
# shapiro-wilk
residuals(he.lands.fit) %>% shapiro.test()

# Bartlett
ncvTest(he.lands.fit)

# Cook's distance
cooks.distance(he.lands.fit) %>% plot()
which(cooks.distance(he.lands.fit) > 4/length(div.df$pop))
```

### EXPORT-- plot (landscape)
```{r}
p1.He <- 
ggplot(div.df, aes(x = pc1_lands , y = He)) + geom_point(size = 5) +
       geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) + 
       #annotate("text", x = 2, y = 0.13, label = "p = 0.08", hjust = 0, size = 8) +
       #annotate("text", x = 2, y = 0.13, label = "italic(R[adj])^2 == 0.13",
       #         parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
       theme_classic() +
       xlab("PC1") + ylab(expression(Expected~heterozygosity~(H[E]))) + ggtitle("(A)") +
       theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20), 
             plot.title = element_text(size = 25, face = "bold"))

print(p1.He)
```

### linear model (local)
```{r}
he.local.fit <- lm(He ~ pc1_local, data = div.df)
summary(he.local.fit)
```

-   Diagnostics
```{r}
plot(he.local.fit)
```

- normality & homoscendasticity
```{r}
# shapiro-wilk
residuals(he.local.fit) %>% shapiro.test()

# Bartlett
ncvTest(he.local.fit)

# Cook's distance
cooks.distance(he.local.fit) %>% plot()
which(cooks.distance(he.local.fit) > 4/length(div.df$pop))
```

### EXPORT-- plot (local)
```{r}
p2.He <- 
ggplot(div.df, aes(x = pc1_local , y = He)) + geom_point(size = 5) +
       geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) + 
       #annotate("text", x = 2, y = 0.13, label = "p = 0.08", hjust = 0, size = 8) +
       #annotate("text", x = 2, y = 0.13, label = "italic(R[adj])^2 == 0.13",
       #         parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
       theme_classic() +
       xlab("PC1") + ylab(expression(Expected~heterozygosity~(H[E]))) + ggtitle("(A)") +
       theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20), 
             plot.title = element_text(size = 25, face = "bold"))

print(p2.He)
```

## Ho ~ PC1
### linear model (landscape)
```{r}
ho.lands.fit <- lm(Ho ~ pc1_lands, data = div.df)
summary(ho.lands.fit)
```

-   Diagnostics
```{r}
plot(ho.lands.fit)
```

- normality & homoscendasticity
```{r}
# shapiro-wilk
residuals(ho.lands.fit) %>% shapiro.test()

# Bartlett
ncvTest(ho.lands.fit)

# Cook's distance
cooks.distance(ho.lands.fit) %>% plot()
which(cooks.distance(ho.lands.fit) > 4/length(div.df$pop))  # 找出 Cook's Distance 較大的觀測值
```

### EXPORT-- plot (landscape)
```{r}
p3.Ho <- 
ggplot(div.df, aes(x = pc1_lands , y = Ho)) + geom_point(size = 5) + 
       geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) + 
       #annotate("text", x = 2, y = 0.13, label = "p = 0.001", hjust = 0, size = 8) +
       #annotate("text", x = 2, y = 0.13, label = "italic(R[adj])^2 == 0.48",
       #         parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
       theme_classic() +
       xlab("PC1") + ylab(expression(Observed~heterozygosity~(H[O]))) + ggtitle("(c)") +
       theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
             plot.title = element_text(size = 25, face = "bold"))
p3.Ho
```

### linear model (local)
```{r}
ho.local.fit <- lm(Ho ~ pc1_local, data = div.df)
summary(ho.local.fit)
```

-   Diagnostics
```{r}
plot(ho.local.fit)
```

- normality & homoscendasticity
```{r}
# shapiro-wilk
residuals(ho.local.fit) %>% shapiro.test()

# Bartlett
ncvTest(ho.local.fit)

# Cook's distance
cooks.distance(ho.local.fit) %>% plot()
which(cooks.distance(ho.local.fit) > 4/length(div.df$pop))
```

### EXPORT-- plot (local)
```{r}
p4.Ho <- 
ggplot(div.df, aes(x = pc1_local , y = Ho)) + geom_point(size = 5) + 
       geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) + 
       #annotate("text", x = 1.5, y = 0.115, label = "p = 0.04", hjust = 0, size = 8) +
       #annotate("text", x = 1.5, y = 0.115, label = "italic(R[adj])^2 == 0.20",
       #         parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
       theme_classic() +
       xlab("PC1") + ylab(expression(Observed~heterozygosity~(H[O]))) + ggtitle("(D)") +
       theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
             plot.title = element_text(size = 25, face = "bold"))
p4.Ho
```

## Ar ~ PC1
### linear model (landscape)
```{r}
ar.lands.fit <- lm(Ar ~ pc1_lands, data = div.df)
summary(ar.lands.fit)
```

-   Diagnostics
```{r}
plot(ar.lands.fit)
```

- normality & homoscendasticity
```{r}
# shapiro-wilk
residuals(ar.lands.fit) %>% shapiro.test()

# Bartlett
ncvTest(ar.lands.fit)

# Cook's distance
cooks.distance(ar.lands.fit) %>% plot()
which(cooks.distance(ar.lands.fit) > 4/length(div.df$pop))
```

### EXPORT-- plot (landscape)
```{r}
p5.Ar <- 
ggplot(div.df, aes(x = pc1_lands , y = Ar)) + geom_point(size = 5) + 
       geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) + 
       #annotate("text", x = 2, y = 1.13, label = "p = 0.005", hjust = 0, size = 8) +
       #annotate("text", x = 2, y = 1.13, label = "italic(R[adj])^2 == 0.39",
       #         parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
       theme_classic() +
       xlab("PC1") + ylab(expression(Allelic~richness~(Ar))) + ggtitle("(E)") +
       theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
             plot.title = element_text(size = 25, face = "bold"))

print(p5.Ar)
```

### linear model (local)
```{r}
ar.local.fit <- lm(Ar ~ pc1_local, data = div.df)
summary(ar.local.fit)
```

-   Diagnostics
```{r}
plot(ar.local.fit)
```

- normality & homoscendasticity
```{r}
# shapiro-wilk
residuals(ar.local.fit) %>% shapiro.test()

# Bartlett
ncvTest(ar.local.fit)

# Cook's distance
cooks.distance(ar.local.fit) %>% plot()
which(cooks.distance(ar.local.fit) > 4/length(div.df$pop))
```

### EXPORT-- plot (local)
```{r}
p6.Ar <- 
ggplot(div.df, aes(x = pc1_local, y = Ar)) + geom_point(size = 5) + 
       geom_smooth(method = "lm", formula = y ~ x, color = "cornflowerblue", se = TRUE) + 
       #annotate("text", x = 1.6, y = 1.125, label = "p = 0.09", hjust = 0, size = 8) +
       #annotate("text", x = 1.6, y = 1.125, label = "italic(R[adj])^2 == 0.12",
       #         parse = TRUE, hjust = 0, vjust = 1.5, size = 8) +
       theme_classic() +
       xlab("PC1") + ylab(expression(Allelic~richness~(Ar))) + ggtitle("(F)") +
       theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20),
             plot.title = element_text(size = 25, face = "bold"))

print(p6.Ar)
```

## EXPORT -- He, Ho, Ar
```{r}
fig4 <- grid.arrange(p1.He, p2.He, p3.Ho, p4.Ho, p5.Ar, p6.Ar, ncol = 2)
```

- tif
```{r eval=FALSE}
ggsave(filename = "Figure4.tif", plot = fig4, width = 18, height = 18, units = "in", dpi = 150)
```

----------------------------------------------------------------------------------------------------
# Diversity & Land use
## import landuse data
```{r}
landuse.lands <- read.csv("../Data/Landuse1000.csv")
landuse.local <- read.csv("../Data/Landuse250.csv")
```

- convert m^2 to km^2
```{r}
landuse.lands <- landuse.lands %>% mutate(across(2:9, ~ .x / 1000000))
landuse.local <- landuse.local %>% mutate(across(2:9, ~ .x / 1000000))

# check area
landuse.lands[ ,2:9] %>% rowSums()
landuse.local[ ,2:9] %>% rowSums()
```

- combine land use & diversity data

```{r}
landuse.lands <- landuse.lands %>% merge(div.df[ ,c(1,7:9)], by = "pop")
landuse.local <- landuse.local %>% merge(div.df[ ,c(1,7:9)], by = "pop")
landuse.lands; landuse.local
```
## Excepted Heterozygosity (landscape)
### model selection
- full model
```{r}
he.lands.full <- lm(He ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.lands)
summary(he.lands.full)
```

```{r}
options(na.action = "na.fail")
dredge(he.lands.full) %>% subset(delta < 1)
```

-   Best model

```{r}
dredge(he.lands.full) %>% get.models(1) %>% .[[1]] %>% summary()
```

## Excepted Heterozygosity (local)
### model selection
- full model
```{r}
he.local.full <- lm(He ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.local)
summary(he.local.full)
```

```{r}
options(na.action = "na.fail")
dredge(he.local.full) %>% subset(delta < 1)
```

-   Best model
```{r}
dredge(he.local.full) %>% get.models(1) %>% .[[1]] %>% summary()
```

## Observed Heterozygosity (landscape)
### model selection
- full model
```{r}
ho.lands.full <- lm(Ho ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.lands)
summary(ho.lands.full)
```

```{r}
options(na.action = "na.fail")
dredge(ho.lands.full) %>% subset(delta < 1)
```

-   Best model
```{r}
dredge(ho.lands.full) %>% get.models(1) %>% .[[1]] %>% summary()
```

## Observed Heterozygosity (local)
### model selection
- full model
```{r}
ho.local.full <- lm(Ho ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.local)
summary(ho.local.full)
```

```{r}
options(na.action = "na.fail")
dredge(ho.local.full) %>% subset(delta < 1)
```

-   Best model
```{r}
dredge(ho.local.full) %>% get.models(1) %>% .[[1]] %>% summary()
```

## Allelic richness (landscape)
### model selection
- full model
```{r}
ar.lands.full <- lm(Ar ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.lands)
summary(ar.lands.full)
```

```{r}
options(na.action = "na.fail")
dredge(ar.lands.full) %>% subset(delta < 1)
```

-   Best model
```{r}
dredge(ar.lands.full) %>% get.models(1) %>% .[[1]] %>% summary()
```

## Allelic richness (local)
### model selection
- full model
```{r}
Ar.local.full <- lm(Ar ~ Agriculture + Forest + Travel + Water +
                    Build + Park + Grass + Other, data = landuse.local)
summary(Ar.local.full)
```

```{r}
options(na.action = "na.fail")
dredge(Ar.local.full) %>% subset(delta < 1)
```

-   Best model
```{r}
dredge(Ar.local.full) %>% get.models(1) %>% .[[1]] %>% summary()
```


----------------------------------------------------------------------------------------------------
# Genetic differentiation (FST)
```{r eval=FALSE}
pfst <- Fst(inputFile = "../Data/snp.gen.txt", outputFile = "", pairs = TRUE)
```

```{r eval=FALSE}
pfst %>% reorder_mat(order = levels(factor(pop))) %>% 
  as.data.frame() %>% write.csv(file = "Pairwise Fst.csv")
```

- read file
```{r}
fst.mat <- read.csv("../Data/Pairwise Fst.csv", row.names = 1, check.names = FALSE)
fst.mat
```

## convert Fst matrix to three columns
```{r}
upper_tri <- upper.tri(fst.mat, diag = FALSE)

fst.df <- data.frame(pop1 = rownames(fst.mat)[row(fst.mat)[upper_tri]],
                     pop2 = colnames(fst.mat)[col(fst.mat)[upper_tri]],
                     Fst  = fst.mat[upper_tri],row.names = NULL)

fst.df
```


-   urbanization region
```{r}
fst.df <- fst.df %>% mutate(region1 = case_when(
                            pop1 %in% c('FLP','NMNS','THU','DFP','DHP','ZGZ') ~ "Urban",
                            pop1 %in% c('BTP','XLC','DKP','DTZ','PLS','DK7-8') ~ "Suburban",
                            pop1 %in% c('WFC','JJF','CLS','DK1-4','BK') ~ "Nonurban")) %>% 

                     mutate(region2 = case_when(
                            pop2 %in% c('FLP','NMNS','THU','DFP','DHP','ZGZ') ~ "Urban",
                            pop2 %in% c('BTP','XLC','DKP','DTZ','PLS','DK7-8') ~ "Suburban",
                            pop2 %in% c('WFC','JJF','CLS','DK1-4','BK') ~ "Nonurban"))

head(fst.df)
```

```{r}
fst.df$pair <- paste(fst.df$region1, fst.df$region2, sep = "–")
levels(as.factor(fst.df$pair))
```

-   some pairs are identical
```{r}
fst.df$pair <- ifelse(fst.df$pair == "Nonurban–Urban", "Urban–Nonurban",
                      ifelse(fst.df$pair == "Nonurban–Suburban", "Suburban–Nonurban",
                             ifelse(fst.df$pair == "Suburban–Urban", "Urban–Suburban", fst.df$pair)))
```

-   reorder pairs
```{r}
fst.df$pair <- factor(fst.df$pair, levels = c("Urban–Urban","Suburban–Suburban","Nonurban–Nonurban",
                                              "Urban–Suburban","Urban–Nonurban","Suburban–Nonurban"))
levels(as.factor(fst.df$pair))    # check levels
```

## statistics
```{r}
# kruskal test
fst.df[fst.df$pair %in% c("Urban–Urban","Suburban–Suburban","Nonurban–Nonurban"),] %>% 
   kruskal.test(Fst ~ pair, data = .)
```

-   Dunn's test
```{r}
fst.df[fst.df$pair %in% c("Urban–Urban","Suburban–Suburban","Nonurban–Nonurban"),] %>% 
  dunn_test(Fst ~ pair, p.adjust.method = "bonferroni")
```

## summarize
```{r}
fst.sum <- Summarize(Fst ~ pair, data = fst.df)
fst.sum
```

## EXPORT (dot plot)
```{r}
fst.plot <- fst.sum[c(1:3),] %>% mutate(pair = c("Urban", "Suburban", "Non-urban"))
fst.plot$pair <- factor(fst.plot$pair, levels = c("Urban", "Suburban", "Non-urban"))
fst.plot
```

- dot plot
```{r}
p.fst <- 
ggplot(data = fst.plot) + 
       geom_pointrange(aes(x = pair, y = mean, ymin = mean-sd/sqrt(n), ymax = mean+sd/sqrt(n),
                           color = pair), size = 1.5) +
       scale_color_manual(values = c("hotpink2","cornflowerblue","mediumseagreen")) +
       labs(x = "", color = "Pair") +  ylab(expression(Pairwise~~F[ST])) +
       theme_classic() +
       theme(axis.title = element_text(size = 15), axis.text.x = element_text(size = 15),
             plot.title = element_text(size = 20, face = "bold")) + 
       theme(legend.position = "none") +
       scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
       annotate("segment", x = 1, xend = 3, y = 0.175, yend = 0.175) +
       annotate("text", x = 2, y = 0.165, label = "****", size = 10) +
       annotate("segment", x = 2, xend = 3, y = 0.15, yend = 0.15) +
       annotate("text", x = 2.5, y = 0.14, label = "**", size = 10) +
       ggtitle("(B)")
p.fst
```



